---
title: "PSTAT 100 Final Project - Fertility and Infant Mortality 2003 - Present"
author: "Arthur Kim and Wesley Kim"
date: "March 8th, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=F, warning=F, message=F}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
library(stats)
library(patchwork)
```

# Abstract
This project examines fertility and infant mortality trends in the United States using two datasets spanning from 2003 to 2023. The analysis focuses on understanding how socioeconomic and healthcare-related factors - specifically the mother's education, Medicaid coverage, and maternal age - are associated with infant mortality rates over time. Key variables explored include the proportion of births to mothers with college degrees, proportion of births covered by Medicaid, and the age distribution of mothers at childbirth. We analyzed how these demographic factors correlate with overall infant mortality rates (IMR), revealing potential patterns and disparities in maternal and infant health outcomes. This report aims to offer insights that can help inform health policymakers and promote equity in maternal and child health in the US.

## Question of Interest

Throughout this report, we will be diving into the question:

# What socioeconomic factors are most predictive of high infant mortality rates within the States?

## Sub-Questions
- Sub-question 1: What role does education of the mother play in predicting infant mortality rates?
- Sub-question 2: How does medicaid coverage influence overall infant mortality rates across states?
- Sub-question 3: What significance does age of the mother play when compared to infant mortality rates?

## Variable Descriptions

- state: Which U.S. state infant statistic was taken
- year: Calender year of the data
- bmcode: Bimonthly period code (1 = Jan-Feb, 2 = Mar-Apr,...)
- bacode: Biannual code (1 = First half of the year, 2 = Second half of the year)
- time: Time period indicator
- start_date: Starting date of the observation period, typically the same as the time variable
- end_date: Ending date of the observation period

## Population Statistics 
- pop_total: Total population in the area
- pop_nhblack: population of Non-Hispanic Black women age 15-44
- pop_nhwhite: population Non-Hispanic White women age 15-44
- pop_hisp: population of Hispanic women age 15-54 population
- pop_otherraceeth: Other races/ethnicities women age 15-54 population
- births_nhblack: Births to non-Hispanic Black mothers
- births_nhwhite: Births to non-Hispanic White mothers
- births_hisp: Births to Hispanic mothers
- births_otherraceeth: Births to mothers of other races/ethnicities
- births_total: Total births across all categories

## By Insurance:

- pop_medicaid: Population covered by Medicaid (women age 15-54 population)
- pop_nonmedicaid: Population not covered by Medicaid (women age 15-54 population)
- births_medicaid: Births covered by Medicaid
- births_nonmedicaid: Births not covered by Medicaid

## By Education:

- pop_nohs: Population without high school education (women age 15-54 population)
- pop_hs: Population with high school education (women age 15-54 population)
- pop_somecoll: Population with some college education (women age 15-54 population)
- pop_coll: Population with college degree (women age 15-54 population)
- births_nohs: Births to mothers without high school education
- births_hs: Births to mothers with high school education
- births_somecoll: Births to mothers with some college education
- births_coll: Births to mothers with college degree

## By Marital Status:

- pop_married: Married population (women age 15-54 population)
- pop_unmarried: Unmarried population (women age 15-54 population)
- births_married: Births to married mothers
- births_unmarried: Births to unmarried mothers

## By Age:

- births_age1524: Births to mother ages 15-24
- births_age2534: Births to mother ages 25-34
- births_age3544: Births to mother ages 35-44
- pop_age1524: Population aged 15-24 (women age 15-54 population)
- pop_age2534: Population aged 25-34 (women age 15-54 population)
- pop_age3544: Population aged 35-44 (women age 15-54 population)

## Mortality

- deaths_nhblack: Number of deaths among non-Hispanic Black population
- deaths_nhwhite: Number of deaths among non-Hispanic White population
- deaths_hisp: Number of deaths among Hispanic population of any race
- deaths_otherraceeth: Number of deaths among other racial/ethnic groups
- deaths_con: Number of congenital-related deaths (present at birth)
- deaths_noncon: Number of non-congenital deaths
- deaths_neo: Number of neonatal deaths (within first 28 days of life)
- deaths_total: Total number of deaths across all categories

## Data Cleaning & Preparation

```{r, echo=FALSE, message=FALSE, warning=FALSE}
mortality <- read_csv('data/infant_mortality_biannual_data.csv', show_col_types = FALSE)
fertility <- read_csv('data/fertility_data.csv', show_col_types = FALSE)

fertility_agg <- fertility %>%
  group_by(state, year, group = case_when(
    bmcode %in% c(1, 2, 3) ~ "H1",  # First half of the year
    bmcode %in% c(4, 5, 6) ~ "H2"   # Second half of the year
  )) %>% 
   summarise(
    across(starts_with("births_"), sum, na.rm = TRUE), # Summing birth count variables
    across(starts_with("pop_"), mean, na.rm = TRUE), # Averaging population variables
    .groups = "drop"
  ) %>%
  mutate(bacode = ifelse(group == "H1", 1, 2)) %>%  # Assign bacode after aggregation
  select(-group)  # Remove temporary grouping variable

data <- mortality %>%
  inner_join(fertility_agg, by = c("state", "year", "bacode"), 
             suffix = c("_m", "_f"))

# Remove duplicate columns (those ending in "_f")
data <- data %>%
  select(-ends_with("_f"))

# Rename columns by removing "_m" suffix from mortality dataset
colnames(data) <- str_replace_all(colnames(data), "_m$", "")
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
data <- data %>%
  mutate(
    IMR_nhblack = (deaths_nhblack / births_nhblack) * 1000,
    IMR_nhwhite = (deaths_nhwhite / births_nhwhite) * 1000,
    IMR_hisp = (deaths_hisp / births_hisp) * 1000,
    IMR_otherraceeth = (deaths_otherraceeth / births_otherraceeth) * 1000,
    IMR_total = (deaths_total / births_total) * 1000, 
    FR_nhblack = (births_nhblack / pop_nhblack) * 1000,
    FR_nhwhite = (births_nhwhite / pop_nhwhite) * 1000,
    FR_hisp = (births_hisp / pop_hisp) * 1000,
    FR_otherraceeth = (births_otherraceeth / pop_otherraceeth) * 1000,
    FR_total = (births_total / pop_total) * 1000,
    prop_married = births_married / births_total,
    prop_college = (births_coll + births_somecoll) / births_total,
    prop_medicaid = births_medicaid / births_total,
    pct_births_age1524 = births_age1524 / births_total,
    pct_births_age2534 = births_age2534 / births_total,
    pct_births_age3544 = births_age3544 / births_total
  )

```

Explain what we did to the dataset and why. Explain the new variables we made

## Infant Mortality Rates by Ethnicity

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Pivot IMR data to long format
IMR_long <- data %>%
  pivot_longer(cols = starts_with("IMR_"), 
               names_to = "Ethnicity", 
               values_to = "IMR") %>%
  mutate(Ethnicity = case_when(
    Ethnicity == "IMR_nhblack" ~ "Non-Hispanic Black",
    Ethnicity == "IMR_nhwhite" ~ "Non-Hispanic White",
    Ethnicity == "IMR_hisp" ~ "Hispanic",
    Ethnicity == "IMR_otherraceeth" ~ "Other Races/Ethnicities",
    Ethnicity == "IMR_total" ~ "Total",
    TRUE ~ Ethnicity  # Keeps any unexpected values unchanged
  ))

# Create scatter plot for IMR by Ethnicity
ggplot(IMR_long, aes(x = Ethnicity, y = IMR)) +
  geom_jitter(width = 0.2, alpha = 0.7, color = "blue") +  # Jitter to avoid overlapping points
  labs(title = "Infant Mortality Rates by Ethnicity",
       x = "Ethnicity",
       y = "Infant Mortality Rate (per 1,000 births)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

IMR_summary <- IMR_long %>%
  group_by(Ethnicity) %>%
  summarise(
    Count = n(),
    Mean_IMR = round(mean(IMR, na.rm = TRUE), 2),
    SD_IMR = round(sd(IMR, na.rm = TRUE), 2)
  )
```

When we first look at the data and compare within different ethnic groups we see that generally Non-Hispanic Black mothers have a higher infant mortality rate compared to the other ethnic groups. The other ethnicities have a generally lower IMF. 

Ethnic Group | Mean IMR | Standard Deviation IMR
---|---|---|
Hispanic | 5.80 | 1.41 |  
Non-Hispanic Black | 11.97 | 3.24 |
Non-Hispanic White | 4.72 | 1.16 |
Other Races/Ethnicity | 4.47 | 2.65 |

When looking through the documentation, it states that the mean IMR for the states is around 5.6. Based on that we see that the Hispanic population is only slightly over, the Non-Hispanic White and other races are a good amount below. However we can see that in the numbers, Non-Hispanic Blacks have 11.97 which is significantly higher than the national average but has the highest standard deviation implying that their are extreme ends. This does not mean just because you have a Non-Hispanic black mother, the infant has a higher chance of mortality. This might show that non-Hispanic Black mothers are being effected by other factors such as education, healthcare access, etc that could be making this number higher. Now we will look into these possible factors.

## Proportion of Births by Different Maternal Age Groups vs IMR 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(data, aes(x = pct_births_age1524, y = IMR_total)) +
  geom_point(color = "darkgreen", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Age 15-24", x = "Proportion of Births", y = "IMR") +
  theme_minimal()

# Plot for 25-34
p2 <- ggplot(data, aes(x = pct_births_age2534, y = IMR_total)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Age 25-34", x = "Proportion of Births", y = "IMR") +
  theme_minimal()

# Plot for 35-44
p3 <- ggplot(data, aes(x = pct_births_age3544, y = IMR_total)) +
  geom_point(color = "purple", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Age 35-44", x = "Proportion of Births", y = "IMR") +
  theme_minimal()

p1 / p2 / p3
```
From the graph above we can see that for mothers that are on the younger side, 15-24, has a positive correlation with IMR (Infant Mortality Rate). This suggests that states with higher proportions of births to younger mothers tend to experience higher infant mortality rates. While mothers that are aged 25-34 and 35-44 both have a negative correlation to IMR, indicating that a higher proportion of births in these older maternal agre groups is associated with lower infant mortality rates. This now leads us to the question, what are the factors between older and younger woman that affects the infant mortality rate? This can be attributed to the 15-24 age group having struggles accessing healthcare services, possibly a lower socioeconomic status, higher rates of unintended pregnancies, and potential lack of education pertaining to an infants health and practices.

## Proportion of Births by College-Educated Mothers vs IMR 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Scatter plot: Proportion of College-Educated Women vs. Infant Mortality Rate
ggplot(data, aes(x = prop_college, y = IMR_total)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Relationship Between College Education and Infant Mortality Rate",
       x = "Proportion of Women with College Education",
       y = "Infant Mortality Rate (per 1,000 births)") +
  theme_minimal() +
  theme(text = element_text(size = 14))
```


## Proportion of Births with Medicaid Birth Coverage vs IMR 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Scatterplot of Medicaid coverage vs. Infant Mortality Rate
ggplot(data, aes(x = prop_medicaid, y = IMR_total)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Relationship Between Medicaid Birth Coverage and Infant Mortality Rate",
    x = "Proportion of Medicaid-Covered Births",
    y = "Infant Mortality Rate (per 1,000 births)"
  ) +
  theme_minimal()
```

